Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}

### 2017-03-06 ident_manu.Rnw
commence hacking

\documentclass[preprint]{elsarticle}
\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
\usepackage{lineno}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage{paralist}
\usepackage{bm}
\usepackage{fixltx2e}
% \usepackage[inline]{showlabels}

@@

@@

@@

Standard approaches for uncertainty analysis can be used to begin addressing model complexity issues.  In the most general sense, uncertainty is evaluated relative to the effects of input conditions or the observed data used to calibrate a model, changes in parameter values, or variation in the structural components (i.e., observational, parameter, or structural uncertainty) \citep{Beck87}.  Evaluating parameter uncertainty is by far the most common and simplest means of evaluating model behavior.  Although uncertainty analyses should be integrated throughout model development and application, parameters are more often evaluated post-hoc as a form of `damage control' for further calibration.  This approach is sometimes called inverse modelling where results from sensitivity analyses are used to guide calibration or fit of the developed model to observations (\citealt{Soetaert10}, or confronting models with data, \textit{sensu} \citealt{Hilborn97}).  Parameter sensitivity analysis combined with inverse modelling necessarily involves questions of parameter `identifiability', where only a subset of parameters can be numerically constrained to the data as compared to the entire parameter set.  Redundancies in parameter effects lead to unidentifiable models where optimal solutions may be empirically impossible (i.e., standard algorithms will not converge) or parameter values may be non-unique leading to the right answer for the wrong reason \citep{Kirchner06}. An unidentifiable parameter or parameter set has effects on model output that can be undone or compensated for by alteration of other parameters.  The concept of identifiable parameter subsets is not foreign to hypoxia or eutrophication models  \citep{Omlin01,Estrada10,Mateus15}, although there is a clear need for greater integration of these concepts in model development \citep{Fasham06}.  Moreover, the inclusion of sensitivity and identifiability analyses in model tuning will require the adoption of conservative selection rules for parameters to calibrate given the number of unique combinations of parameter subsets for most models \citep[e.g.,][]{Wagener01, Wagener01b}.

@@

This study evaluated the core unit of a recently developed hydrodynamic and ecological model that describes horizontal and vertical transport and mixing of state variables relevant for hypoxia in the northern \ac{gom}.  The \ac{cgem} includes elements from the Navy Coastal Ocean Model \citep{Martin00} for hydrodynamics on the \ac{lcs} and a biogeochemical model with multiple plankton groups, water-column metabolism, and sediment diagenesis \citep{Eldridge10}.  The hydrodynamic component of \ac{cgem} provides a spatially-explicit description of hypoxia using an orthogonal grid with an approximate horizontal resolution of 1.9 km$^2$ and twenty equally-spaced vertical sigma layers on the shelf (depth $\leq$ 100 m, with additional hybrid layers at deeper depths).  The biogeochemical component includes equations for 36 state variables including six phytoplankton groups (with nitrogen and phosophorus quotas for each), two zooplankton groups, nitrate, ammonium, phosphate, dissolved inorganic carbon, oxygen, silica, and multiple variables for dissolved and particulate organic matter from different sources.  Atmospheric and hydrological boundary conditions described in \citet{Hodur97} and \citet{Lehrter13} are also included in \ac{cgem}.

@@

Specific approaches can be used to improve and build on the results presented herein, in addition to the more general considerations noted above. An evaluation of model precision following calibration with relevant parameter subsets could provide additional information that supports results from the sensitivity analyses.  For simplicity, our analysis did not calibrate model parameters and an explicit assumption was that parameter subsets with $\gamma$ below 10 or 15 were identifiable.  To our knowledge, this threshold has not been rigorously evaluated and it is likely to vary between parameter subsets and the chosen calibration method.  Variation in parameter estimates given the calibration method and different identifiability thresholds could affect the interpration of model output.  Further, the extent to which results for the \ac{zerod} model extrapolate to the larger three-dimensional model should be evaluated.  Although the above analyses were facilitated using the \ac{zerod} model (i.e, quick execution times, ease of changing model parameters), lack of physical transport or spatial components potentially limits extrapolation of the results. Uncertainy in parameter estimates at lower dimensions could be magnified as errors propagate to larger scales or site-specific observations for model calibration are not appropriate across a hydrodynamic grid \citep{Harvey00,Lehrter10}. The effects of structural or observational uncertainty should also be evaluated in the context of the larger model.  Our simple approach to examine changes in the initial conditions and switches used in the model showed that sensitivity estimates of individual parameters differed relative to the original model configuration, having further implications for characterizing identifiability.

>>>>

\documentclass[preprint]{elsarticle}
\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
\usepackage{lineno}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage{paralist}
\usepackage{bm}
\usepackage{fixltx2e}
% \usepackage[inline]{showlabels}

@@

\linespread{1.5}

@@

\author{Brandon M. Jarvis}
\address{USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}
\ead{jarvis.brandon@epa.gov}

@@

Standard approaches for uncertainty analysis can be used to begin addressing model complexity issues.  In the most general sense, uncertainty is evaluated relative to the effects of input conditions or the observed data used to calibrate a model, changes in parameter values, or variation in the structural components (i.e., observational, parameter, or structural uncertainty) \citep{Beck87}.  Evaluating parameter uncertainty is by far the most common and simplest means of evaluating model behavior.  Although uncertainty analyses should be integrated throughout model development and application, parameters are more often evaluated post-hoc as a form of `damage control' for further calibration.  This approach is sometimes called inverse modelling where results from sensitivity analyses are used to guide calibration or fit of the developed model to observations (\citealt{Soetaert10}, or confronting models with data, \textit{sensu} \citealt{Hilborn97}).  Parameter sensitivity analysis combined with inverse modelling necessarily involves questions of parameter `identifiability', where only a subset of identifiable parameters can be numerically constrained to the data as compared to the entire set.  Redundancies in parameter effects lead to unidentifiable models where optimal solutions may be empirically impossible (i.e., standard algorithms will not converge) or parameter values may be non-unique leading to the right answer for the wrong reason \citep{Kirchner06}. An unidentifiable parameter or parameter set has effects on model output that can be undone or compensated for by alteration of other parameters.  The concept of identifiable parameter subsets is not foreign to hypoxia or eutrophication models  \citep{Omlin01,Estrada10,Mateus15}, although there is a clear need for greater integration of these concepts in model development \citep{Fasham06}.  Moreover, the inclusion of sensitivity and identifiability analyses in model tuning will require the adoption of conservative selection rules for parameters to calibrate given the number of unique combinations of parameter subsets for most models \citep[e.g.,][]{Wagener01, Wagener01b}.

@@

This study evaluated the core unit of a recently developed hydrodynamic and ecological model that describes horizontal and vertical transport and mixing of state variables relevant for hypoxia in the northern \ac{gom}.  The \ac{cgem} includes elements from the Navy Coastal Ocean Model \citep{Martin00} for hydrodynamics on the \ac{lcs} and a biogeochemical model with multiple plankton groups, water-column metabolism, and sediment diagenesis \citep{Eldridge10}.  The hydrodynamic component of \ac{cgem} provides a spatially-explicit description of hypoxia using an orthogonal grid with an approximate horizontal resolution of 1.9 km$^2$ and twenty equally-spaced vertical sigma layers on the shelf (depth $\leq$ 100 m, with additional hybrid layers at deeper depths).  The biogeochemical component includes equations for 36 state variables including six phytoplankton groups (with nitrogen and phosophorus quotas for each), two zooplankton groups, nitrate, ammonium, phosphate, dissolved inorganic carbon, oxygen, silica, and multiple variables for dissolved and particulate organic matter from different sources.  Atmospheric and hydrologic boundary conditions described in \citet{Hodur97} and \citet{Lehrter13} are also included in \ac{cgem}.

@@

Specific approaches can be used to improve and build on the results presented herein, in addition to the more general considerations noted above. An evaluation of model precision following calibration with relevant parameter subsets could provide additional information that supports results from the sensitivity analyses.  For simplicity, our analysis did not calibrate model parameters and an explicit assumption was that parameter subsets with $\gamma$ below 10 or 15 were identifiable.  To our knowledge, this threshold has not been rigorously evaluated and it is likely to vary between parameter subsets and the chosen calibration method.  Variation in parameter estimates given the calibration method and different identifiability thresholds could affect the interpration of model output.  Further, the extent to which results for the \ac{zerod} model extrapolate to the larger three-dimensional model should be evaluated.  Although the above analyses were facilitated using the \ac{zerod} model (i.e, quick execution times, ease of changing model parameters), lack of physical transport or spatial components potentially limits extrapolation of the results. Uncertainy in parameter estimates at lower dimensions could be magnified as errors propagate to larger scales or site-specific observations for model calibration are not appropriate across a hydrodynamic grid \citep{Harvey00,Lehrter10}. The effects of structural or observational uncertainty should also be evaluated in the context of the larger model.  Our simple approach to examine changes in the initial conditions and switches used in the model showed that sensitivity estimates of individual parameters differed relative to the original model configuration, having further implications for characterizing identifiability.
\clearpage

### 2017-03-06 ident_manu.tex
commence hacking

\usepackage{alltt}
\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
\usepackage{lineno}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage{paralist}
\usepackage{bm}
\usepackage{fixltx2e}
% \usepackage[inline]{showlabels}

@@

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Fri Dec 16 10:08:47 2016 -0600, \href{https://github.com/fawda123/identifiability/commit/5e43ad8aba3a34ec8163993b8c3ffa785e1f0fcd}{5e43ad8aba3a34ec8163993b8c3ffa785e1f0fcd}}

@@

% initial conditions evaluated 
%latex.default(inits[, -1], file = "", rowlabel = "Variables",     caption = cap_val, caption.loc = "top", label = "tab:inits",     rowname = inits[, 1])%
\begin{table}[!tbp]
\caption{Initial conditions that were varied to evaluate effects of observational uncertainty on parameter sensitivity.  April and September values are based on seasonal averages from water quality samples of the \ac{lcs}.  All units are $\mu$mol L$^{-1}$.\label{tab:inits}} 
\begin{center}
\begin{tabular}{llll}
\hline\hline
\multicolumn{1}{l}{Variables}&\multicolumn{1}{c}{Default}&\multicolumn{1}{c}{April}&\multicolumn{1}{c}{September}\tabularnewline
\hline
Dissolved Inorganic Carbon&$2.13\times 10^{3}$&$2.1\times 10^{3}$&$2.13\times 10^{3}$\tabularnewline
Ammonium&$1.09$&$1.79$&$0.35$\tabularnewline
Nitrate&$71.4$&$3.75$&$2.04$\tabularnewline
Dissolved Oygen&$172$&$194.47$&$152.57$\tabularnewline
Phosphate&$1.81$&$0.23$&$0.42$\tabularnewline
Silica&$71.4$&$5.84$&$9.77$\tabularnewline
\hline
\end{tabular}\end{center}

% structural conditions
%latex.default(swtch[, -1], file = "", rowlabel = "Switch type",     caption = cap_val, caption.loc = "top", label = "tab:strcs",     rowname = swtch[, 1])%
\begin{table}[!tbp]
\caption{Model switches that were used to evaluate effects of structural uncertainty on parameter sensitivity. Complete details and equations are provided as supplementary material to \citetalias{LehrterIP}.\label{tab:strcs}} 
\begin{center}
\begin{tabular}{lll}
\hline\hline
\multicolumn{1}{l}{Switch type}&\multicolumn{1}{c}{Default}&\multicolumn{1}{c}{Complex}\tabularnewline
\hline
Temperature&sigmoidal {\scriptsize \citep{Eldridge10}}&Arrenhius {\scriptsize \citep{Geider97}}\tabularnewline
Uptake&Michaelis-Menten {\scriptsize \citep{Dugdale67}}&Geider {\scriptsize \citep{Lehman75,Geider98}}\tabularnewline
Quota&Droop {\scriptsize \citep{Droop73}}&Flynn {\scriptsize \citep{Flynn03}}\tabularnewline
Chla:C&regression {\scriptsize \citep{Murrell14}}&Cloern {\scriptsize \citep{Cloern95}}\tabularnewline
Photosynthesis&photoinhibition {\scriptsize \citep{Platt80}}&nutrient dependent\tabularnewline
Specific growth rate&Leibig's minimum&nutrient dependent\tabularnewline
\hline
\end{tabular}\end{center}

>>>>

\usepackage{alltt}
\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
\usepackage{lineno}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage{paralist}
\usepackage{bm}
\usepackage{fixltx2e}
% \usepackage[inline]{showlabels}

@@

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Wed Dec 21 10:59:02 2016 -0600, \href{https://github.com/fawda123/identifiability/commit/d3eb2370a7cc79abbb571fc9216ed07baa4c881b}{d3eb2370a7cc79abbb571fc9216ed07baa4c881b}}

@@

% initial conditions evaluated 
%latex.default(inits[, -1], file = "", rowlabel = "Variables",     caption = cap_val, caption.loc = "top", label = "tab:inits",     rowname = inits[, 1])%
\begin{table}[!tbp]
\caption{Initial conditions that were varied to evaluate effects of observational uncertainty on parameter sensitivity.  April and September values are based on seasonal averages from water quality samples of the \ac{lcs}.  All units are $\mu$mol L$^{-1}$.\label{tab:inits}} 
\begin{center}
\begin{tabular}{llll}
\hline\hline
\multicolumn{1}{l}{Variables}&\multicolumn{1}{c}{Default}&\multicolumn{1}{c}{April}&\multicolumn{1}{c}{September}\tabularnewline
\hline
Dissolved Inorganic Carbon&$2.13\times 10^{3}$&$2.1\times 10^{3}$&$2.13\times 10^{3}$\tabularnewline
Ammonium&$1.09$&$1.79$&$0.35$\tabularnewline
Nitrate&$71.4$&$3.75$&$2.04$\tabularnewline
Dissolved Oygen&$172$&$194.47$&$152.57$\tabularnewline
Phosphate&$1.81$&$0.23$&$0.42$\tabularnewline
Silica&$71.4$&$5.84$&$9.77$\tabularnewline
\hline
\end{tabular}\end{center}
\end{table}

% structural conditions
%latex.default(swtch[, -1], file = "", rowlabel = "Switch type",     caption = cap_val, caption.loc = "top", label = "tab:strcs",     rowname = swtch[, 1])%
\begin{table}[!tbp]
\caption{Model switches that were used to evaluate effects of structural uncertainty on parameter sensitivity. Complete details and equations are provided as supplementary material to \citetalias{LehrterIP}.\label{tab:strcs}} 
\begin{center}
\begin{tabular}{lll}
\hline\hline
\multicolumn{1}{l}{Switch type}&\multicolumn{1}{c}{Default}&\multicolumn{1}{c}{Complex}\tabularnewline
\hline
Temperature&sigmoidal {\scriptsize \citep{Eldridge10}}&Arrenhius {\scriptsize \citep{Geider97}}\tabularnewline
Uptake&Michaelis-Menten {\scriptsize \citep{Dugdale67}}&Geider {\scriptsize \citep{Lehman75,Geider98}}\tabularnewline
Quota&Droop {\scriptsize \citep{Droop73}}&Flynn {\scriptsize \citep{Flynn03}}\tabularnewline
Chla:C&regression {\scriptsize \citep{Murrell14}}&Cloern {\scriptsize \citep{Cloern95}}\tabularnewline
Photosynthesis&photoinhibition {\scriptsize \citep{Platt80}}&nutrient dependent\tabularnewline
Specific growth rate&Leibig's minimum&nutrient dependent\tabularnewline
\hline
\end{tabular}\end{center}
\end{table}
